---
title: "Predicting Breast Cancer Mortality from Clinical and Genomic Data (METABRIC)"
author: "Saba Naji"
output:
  github_document:
    toc: true
    html_preview: false
---

```{r knitr-options, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.path = "figures/",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 110
)
```

```{r setup}
# Data handling and plots
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

# Modelling
library(caret)      # training / cross-validation framework
library(glmnet)     # Lasso and Ridge logistic regression
library(pROC)       # ROC curves and AUC
library(kernlab)    # SVM with RBF kernel (used by caret "svmRadial")
library(rpart)      # decision trees
library(ranger)     # random forest
library(gbm)        # boosting with the AdaBoost (exponential) loss
library(doParallel) # parallel cross-validation

set.seed(123)
```

# Part 1. Data Exploration and Feature Selection

## 1.1 Loading the data

```{r load-data}
data_path <- file.path("..", "data", "METABRIC_RNA_Mutation.csv")

# guess_max is large so that mutation columns (mostly "0" in the first rows,
# protein-change labels such as "H1047R" later) are read as text, not numbers
df_raw <- read_csv(data_path, guess_max = 5000, show_col_types = FALSE)

# Three column names contain "-" or "+" or start with a digit, which breaks
# R formulas and some models (rpart). They are renamed to syntactic names.
df_raw <- df_raw |>
  rename(
    pam50_claudin_low_subtype     = `pam50_+_claudin-low_subtype`,
    three_gene_classifier_subtype = `3-gene_classifier_subtype`,
    hla_g                         = `hla-g`
  )

dim(df_raw)
```

The dataset has `r nrow(df_raw)` patients and `r ncol(df_raw)` columns. The
columns fall into three groups:

* **clinical** variables (columns 1–31): age, tumour size, grade, receptor
  status, treatments, survival outcome, …
* **gene expression** (mRNA z-scores) for 489 genes (columns 32–520),
* **mutation** columns for 173 genes (names ending in `_mut`). Each cell holds
  `0` when the gene is not mutated, or the protein change (for example
  `H1047R`) when it is.

```{r feature-groups-by-position}
# Keep the original column groups so that we can build feature subsets later
clinical_cols   <- names(df_raw)[1:31]
mutation_cols   <- grep("_mut$", names(df_raw), value = TRUE)
expression_cols <- setdiff(names(df_raw)[32:ncol(df_raw)], mutation_cols)

c(clinical = length(clinical_cols),
  expression = length(expression_cols),
  mutation = length(mutation_cols))
```

## 1.2 Mutation columns as binary indicators

In the raw file a mutation column can have hundreds of different values (one per
protein change). Treating them as factors would create thousands of dummy
variables, most of them seen in only one patient. For prediction the useful
information is whether the gene is mutated or not, so each mutation column is
recoded to 0/1.

```{r binary-mutations}
n_levels_before <- sapply(df_raw[mutation_cols], n_distinct)
summary(n_levels_before)

df_raw <- df_raw |>
  mutate(across(all_of(mutation_cols), ~ as.integer(.x != "0")))

# Number of patients with a mutation in the five most frequently mutated genes
sort(colSums(df_raw[mutation_cols]), decreasing = TRUE)[1:5]
```

## 1.3 Defining the target

```{r define-target}
table(df_raw$death_from_cancer, useNA = "ifany")

df <- df_raw |>
  filter(death_from_cancer %in% c("Living", "Died of Disease")) |>
  mutate(
    death_from_cancer = factor(
      if_else(death_from_cancer == "Died of Disease", "Died", "Living"),
      levels = c("Died", "Living")
    )
  )

table(df$death_from_cancer)
```

Only patients who are **Living** or **Died of Disease** are kept. Patients who
died of other causes (`r sum(df_raw$death_from_cancer == "Died of Other Causes", na.rm = TRUE)`)
and one patient with a missing outcome are removed, so the target represents
breast-cancer–specific death only. This leaves `r nrow(df)` patients
(`r sum(df$death_from_cancer == "Died")` Died, `r sum(df$death_from_cancer == "Living")`
Living). The positive class for all metrics is **Died**.

## 1.4 Removing identifiers and leakage variables

```{r leakage}
# overall_survival is identical to the target after the filter above
table(df$overall_survival, df$death_from_cancer)

leakage_cols <- c("patient_id", "overall_survival", "overall_survival_months")
df <- df |> select(-all_of(leakage_cols))
```

Three columns must not be used as predictors:

* `patient_id` is only an identifier.
* `overall_survival` is exactly the same as the target once the other causes of
  death are removed (see the table above).
* `overall_survival_months` is the follow-up time. It is measured at the end of
  the study, not at diagnosis, and patients who died of cancer have much shorter
  follow-up by definition. Using it would leak the outcome into the model.

## 1.5 Numeric variables that are really categorical

```{r categorical-numeric}
num_cols <- names(df)[sapply(df, is.numeric)]
num_cols <- setdiff(num_cols, mutation_cols)

# Numeric columns with 10 or fewer distinct values
cat_like_numeric <- num_cols[sapply(df[num_cols], function(x) n_distinct(x, na.rm = TRUE) <= 10)]
cat_like_numeric

df <- df |>
  mutate(across(all_of(cat_like_numeric), as.factor)) |>
  mutate(across(where(is.character), as.factor))
```

Treatment indicators (`chemotherapy`, `hormone_therapy`, `radio_therapy`),
`cohort` and `neoplasm_histologic_grade` are stored as numbers but describe
groups, so they are converted to factors. All remaining text columns are
converted to factors as well. Mutation indicators stay as 0/1 numbers.

```{r summary-clinical}
summary(df[, intersect(clinical_cols, names(df))])
```

## 1.6 Missing values

```{r missingness}
na_count <- colSums(is.na(df))
summary(na_count)
sort(na_count[na_count > 0], decreasing = TRUE)
```

Missing values are found only in clinical columns; gene expression and
mutation columns are complete. `tumor_stage`
(`r na_count[["tumor_stage"]]` missing, `r round(100 * mean(is.na(df$tumor_stage)))`%)
and `three_gene_classifier_subtype`
(`r na_count[["three_gene_classifier_subtype"]]` missing) have the most
missing values, so they are removed instead of being heavily imputed. The other
missing values are imputed later using the training set only (Section 1.9).

```{r drop-high-missing}
cols_to_drop <- c("tumor_stage", "three_gene_classifier_subtype")
df <- df |> select(-all_of(cols_to_drop))
dim(df)
```

## 1.7 Exploratory plots

```{r eda-plots, fig.height = 3.5}
eda_vars <- c("age_at_diagnosis", "tumor_size", "lymph_nodes_examined_positive",
              "nottingham_prognostic_index", "chemotherapy", "radio_therapy",
              "hormone_therapy", "er_status", "pr_status", "her2_status")

for (v in eda_vars) {
  plot_data <- df |> filter(!is.na(.data[[v]]))
  if (is.numeric(df[[v]])) {
    p <- ggplot(plot_data, aes(x = .data[[v]], fill = death_from_cancer)) +
      geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
      labs(title = paste("Distribution of", v, "by outcome"), fill = "Outcome")
  } else {
    p <- ggplot(plot_data, aes(x = .data[[v]], fill = death_from_cancer)) +
      geom_bar(position = "fill") +
      labs(title = paste("Proportion of outcome by", v), y = "Proportion", fill = "Outcome")
  }
  print(p + theme_minimal())
}
```

Age at diagnosis shows only a small shift between the two groups. Tumour size
and the number of positive lymph nodes separate the groups more clearly: large
tumours and many positive nodes are more common among patients who died of the
disease. The Nottingham Prognostic Index (NPI), which combines size, nodes and
grade, gives the clearest separation.

Chemotherapy is more common among patients who died, which most likely reflects
that chemotherapy is given to more severe cases (confounding by indication), not
a harmful treatment effect. ER-positive and PR-positive patients have a lower
proportion of cancer deaths, while HER2-positive patients have a higher one.
Overall, disease severity (NPI, size, nodes) and receptor status look like the
most useful clinical predictors.

## 1.8 Train / test split

All preprocessing steps that learn something from the data (imputation values,
correlation filter, scaling) are estimated on the training set only and then
applied to the test set. This keeps the test set completely unseen.

```{r split}
train_index <- createDataPartition(df$death_from_cancer, p = 0.8, list = FALSE)
train_raw <- df[train_index, ]
test_raw  <- df[-train_index, ]

rbind(train = table(train_raw$death_from_cancer),
      test  = table(test_raw$death_from_cancer))
```

The split is stratified by the outcome, so both sets have almost the same
proportion of deaths (`r round(100 * mean(train_raw$death_from_cancer == "Died"), 1)`%
in training, `r round(100 * mean(test_raw$death_from_cancer == "Died"), 1)`% in test).

## 1.9 Imputation (training statistics only)

```{r imputation}
Mode <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Median for numeric columns, mode for factors.
# The fill values are computed on `reference` (the training set).
impute_mixed <- function(data, reference) {
  out <- data
  for (nm in names(out)) {
    if (!anyNA(out[[nm]])) next
    if (is.factor(out[[nm]])) {
      out[[nm]][is.na(out[[nm]])] <- Mode(reference[[nm]])
    } else if (is.numeric(out[[nm]])) {
      out[[nm]][is.na(out[[nm]])] <- median(reference[[nm]], na.rm = TRUE)
    }
  }
  out
}

train_imp <- impute_mixed(train_raw, reference = train_raw)
test_imp  <- impute_mixed(test_raw,  reference = train_raw)

c(train_missing = sum(is.na(train_imp)), test_missing = sum(is.na(test_imp)))
```

## 1.10 Removing uninformative and redundant features

```{r uninformative-features}
# (a) factors with a single level in the training set
single_level <- names(train_imp)[sapply(train_imp, function(x) is.factor(x) && nlevels(droplevels(x)) < 2)]

# (b) very rare mutations: fewer than 10 mutated patients in the training set
mut_in_data <- intersect(mutation_cols, names(train_imp))
rare_mut <- mut_in_data[colSums(train_imp[mut_in_data]) < 10]

single_level
length(rare_mut)
```

`r if (length(single_level) == 0) "No factor has a single level in the training data, but" else paste(length(single_level), "factor(s) have a single level in the training data, and")`
`r length(rare_mut)` of the `r length(mutation_cols)` mutation indicators are
mutated in fewer than 10 training patients. Such columns
carry almost no information and make cross-validation folds unstable, so they
are removed.

```{r correlation}
train_imp <- train_imp |> select(-all_of(c(single_level, rare_mut)))

num_pred_cols <- setdiff(names(train_imp)[sapply(train_imp, is.numeric)], mutation_cols)
cor_mat <- cor(train_imp[, num_pred_cols])

high_corr <- findCorrelation(cor_mat, cutoff = 0.80, names = TRUE, exact = TRUE)
high_corr

# Pairs with |r| > 0.80, to see which feature each removed one is related to
high_pairs <- which(abs(cor_mat) > 0.80 & upper.tri(cor_mat), arr.ind = TRUE)
data.frame(var1 = rownames(cor_mat)[high_pairs[, 1]],
           var2 = colnames(cor_mat)[high_pairs[, 2]],
           r    = round(cor_mat[high_pairs], 3))
```

Highly correlated predictors make coefficient estimates unstable, especially in
linear models. `findCorrelation()` with a cut-off of 0.80 marks
`r paste0("\x60", high_corr, "\x60", collapse = " and ")` for removal (the
feature of each pair with the larger mean absolute correlation to the rest).
The number of highly correlated pairs is small because the gene expression
values are already normalised and most genes are only weakly correlated.

```{r apply-feature-removal}
removed_cols <- c(single_level, rare_mut, high_corr)

train_red <- train_raw |> select(-any_of(removed_cols))
train_red <- impute_mixed(train_red, reference = train_raw)
test_red  <- impute_mixed(test_raw |> select(-any_of(removed_cols)), reference = train_raw)

dim(train_red)
```

## 1.11 Scaling (training statistics only)

```{r scaling}
scale_cols <- setdiff(names(train_red)[sapply(train_red, is.numeric)], mutation_cols)

scaler <- preProcess(train_red[, scale_cols], method = c("center", "scale"))

train_data <- train_red
test_data  <- test_red
train_data[, scale_cols] <- predict(scaler, train_red[, scale_cols])
test_data[, scale_cols]  <- predict(scaler, test_red[, scale_cols])

# Drop unused factor levels so that train and test dummies match
train_data <- droplevels(train_data)
for (nm in names(test_data)) {
  if (is.factor(test_data[[nm]])) {
    test_data[[nm]] <- factor(test_data[[nm]], levels = levels(train_data[[nm]]))
  }
}
test_data <- impute_mixed(test_data, reference = train_data)  # levels not seen in training -> mode
```

All continuous features are centred and scaled with the mean and standard
deviation of the training set, so that features on different scales contribute
equally (important for penalised regression, KNN and SVM). Mutation indicators
stay 0/1.

## 1.12 PCA

```{r pca}
pca_cols <- setdiff(names(train_data)[sapply(train_data, is.numeric)], mutation_cols)
pca_fit <- prcomp(train_data[, pca_cols], center = FALSE, scale. = FALSE)

var_explained <- pca_fit$sdev^2 / sum(pca_fit$sdev^2)
round(100 * cumsum(var_explained)[c(2, 3, 10, 50)], 1)

pca_vis <- data.frame(pca_fit$x[, 1:3], outcome = train_data$death_from_cancer)

ggplot(pca_vis, aes(PC1, PC2, color = outcome)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(title = "PCA of training data: PC1 vs PC2", color = "Outcome")

ggplot(pca_vis, aes(PC1, PC3, color = outcome)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(title = "PCA of training data: PC1 vs PC3", color = "Outcome")
```

The first two components explain only `r round(100 * sum(var_explained[1:2]), 1)`%
of the variance (the first three `r round(100 * sum(var_explained[1:3]), 1)`%),
and the two classes overlap strongly in all projections. The outcome signal is
therefore spread over many dimensions or is non-linear, so a simple 2-D or 3-D
linear projection is not enough to separate the classes. Classification models
that use all features are needed.

## 1.13 Summary of the preprocessing

```{r dims}
data.frame(
  step = c("raw data", "after target filter", "after dropping leakage / high-NA columns",
           "training set (final)", "test set (final)"),
  rows = c(nrow(df_raw), nrow(df_raw |> filter(death_from_cancer %in% c("Living", "Died of Disease"))),
           nrow(df), nrow(train_data), nrow(test_data)),
  columns = c(ncol(df_raw), ncol(df_raw), ncol(df), ncol(train_data), ncol(test_data))
)
```

The number of patients drops from `r nrow(df_raw)` to `r nrow(df)` only because
of the target definition (Section 1.3); no patient is removed for missing
values. The final modelling data has `r ncol(train_data) - 1` predictors.

# Part 2. Baseline Models: Lasso and Ridge Logistic Regression

## 2.1 Feature subsets

```{r feature-subsets}
all_features <- setdiff(names(train_data), "death_from_cancer")

mutation_features   <- intersect(mutation_cols, all_features)
expression_features <- intersect(expression_cols, all_features)

genetic_features     <- c(expression_features, mutation_features, "mutation_count")
non_genetic_features <- setdiff(all_features, genetic_features)
combined_features    <- all_features

c(non_genetic = length(non_genetic_features),
  genetic     = length(genetic_features),
  combined    = length(combined_features))
```

* **Non-genetic**: clinical, pathological and treatment variables.
* **Genetic**: gene expression, mutation indicators and `mutation_count`.
* **Combined**: all features.

## 2.2 Cross-validation set-up and helper functions

```{r train-control}
train_control <- trainControl(
  method = "cv", number = 5,
  savePredictions = "final",
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

f1_score <- function(p, r) {
  if (is.na(p) || is.na(r) || (p + r) == 0) NA_real_ else 2 * p * r / (p + r)
}

# Metrics from a confusion matrix and predicted probabilities
metric_row <- function(pred_class, prob_died, truth, positive = "Died") {
  cm <- confusionMatrix(pred_class, truth, positive = positive)
  roc_obj <- roc(truth, prob_died, levels = c("Living", "Died"), direction = "<", quiet = TRUE)
  precision <- as.numeric(cm$byClass["Pos Pred Value"])
  recall    <- as.numeric(cm$byClass["Sensitivity"])
  tibble(
    Accuracy  = as.numeric(cm$overall["Accuracy"]),
    Precision = precision,
    Recall    = recall,
    F1        = f1_score(precision, recall),
    ROC_AUC   = as.numeric(auc(roc_obj))
  )
}

# Test-set metrics for a caret model
eval_on_test <- function(model, test_sub, positive = "Died") {
  metric_row(
    pred_class = predict(model, newdata = test_sub),
    prob_died  = predict(model, newdata = test_sub, type = "prob")[, positive],
    truth      = test_sub$death_from_cancer
  )
}

# Mean and SD of the metrics over the 5 CV folds (best tuning parameters only)
cv_mean_sd <- function(model, positive = "Died") {
  pred_df <- model$pred
  for (nm in names(model$bestTune)) {
    pred_df <- pred_df[pred_df[[nm]] == model$bestTune[[nm]], ]
  }
  pred_df |>
    group_by(Resample) |>
    group_modify(~ metric_row(.x$pred, .x[[positive]], .x$obs)) |>
    ungroup() |>
    summarise(across(c(Accuracy, Precision, Recall, F1, ROC_AUC),
                     list(mean = ~ mean(.x, na.rm = TRUE), sd = ~ sd(.x, na.rm = TRUE))))
}

# Fit a caret model on one feature subset and evaluate it
fit_caret_model <- function(method, grid, feature_set, algorithm, subset_name, ...) {
  train_sub <- train_data[, c(feature_set, "death_from_cancer")]
  test_sub  <- test_data[,  c(feature_set, "death_from_cancer")]

  set.seed(123)  # same CV folds for every model
  model <- train(
    death_from_cancer ~ .,
    data = train_sub,
    method = method,
    trControl = train_control,
    metric = "ROC",
    tuneGrid = grid,
    ...
  )

  list(
    Algorithm = algorithm,
    Subset    = subset_name,
    model     = model,
    bestTune  = model$bestTune,
    CV        = cv_mean_sd(model),
    Test      = eval_on_test(model, test_sub),
    test_prob = predict(model, newdata = test_sub, type = "prob")[, "Died"]
  )
}

subsets <- list(
  "Non-Genetic" = non_genetic_features,
  "Genetic"     = genetic_features,
  "Combined"    = combined_features
)
```

A 5-fold cross-validation on the training set is used to tune the
hyper-parameters, with ROC-AUC as the selection metric. The seed is reset before
each model so that all models see the same folds. The held-out test set is used
only once per model, for the final evaluation.

## 2.3 Lasso and Ridge on the three subsets

```{r lasso-ridge}
lambda_grid <- 10^seq(-4, 1, length.out = 60)

glmnet_results <- list()
for (s in names(subsets)) {
  glmnet_results[[paste("Lasso", s)]] <- fit_caret_model(
    "glmnet", expand.grid(alpha = 1, lambda = lambda_grid), subsets[[s]], "Lasso", s)
  glmnet_results[[paste("Ridge", s)]] <- fit_caret_model(
    "glmnet", expand.grid(alpha = 0, lambda = lambda_grid), subsets[[s]], "Ridge", s)
}
```

`alpha = 1` gives the Lasso penalty (some coefficients become exactly zero) and
`alpha = 0` gives the Ridge penalty (all coefficients are shrunk but kept). The
penalty strength `lambda` is searched on a log scale from 10^-4 to 10, because
Ridge usually needs a much larger `lambda` than Lasso when there are hundreds
of predictors.

```{r results-table-helper}
pack_row <- function(obj) {
  tibble(Algorithm = obj$Algorithm, Subset = obj$Subset) |>
    bind_cols(obj$CV |> rename_with(~ paste0("CV_", .x))) |>
    bind_cols(obj$Test |> rename_with(~ paste0("Test_", .x)))
}

show_table <- function(tbl) {
  tbl |>
    select(Algorithm, Subset, CV_ROC_AUC_mean, CV_ROC_AUC_sd,
           Test_ROC_AUC, Test_Accuracy, Test_Precision, Test_Recall, Test_F1) |>
    mutate(across(where(is.numeric), ~ round(.x, 3))) |>
    knitr::kable()
}
```

```{r glmnet-results}
glmnet_tbl <- bind_rows(lapply(glmnet_results, pack_row)) |> arrange(desc(CV_ROC_AUC_mean))
show_table(glmnet_tbl)

# Chosen penalty for each model
sapply(glmnet_results, function(x) signif(x$bestTune$lambda, 3))
```

```{r glmnet-comment, include=FALSE}
best_lasso_name <- glmnet_tbl |> filter(Algorithm == "Lasso") |> slice(1)
best_ridge_name <- glmnet_tbl |> filter(Algorithm == "Ridge") |> slice(1)
```

The Lasso model with the combined features has the best mean CV ROC-AUC
(`r round(glmnet_tbl$CV_ROC_AUC_mean[1], 3)`) and also the best test ROC-AUC
(`r round(glmnet_tbl$Test_ROC_AUC[1], 3)`). However, the Lasso and Ridge models on
the non-genetic features are very close (CV ROC-AUC about 0.76), and the
difference is smaller than the fold-to-fold standard deviation. Models that use
only the genetic features are clearly weaker (ROC-AUC about 0.70). So most of
the predictive signal comes from the clinical variables, and gene expression and
mutations add only a small amount on top of them.

Ridge needs a much larger penalty than Lasso on the genetic and combined
subsets (lambda around 1 compared with 0.03), because it has to shrink hundreds
of coefficients at the same time instead of setting most of them to zero.

With the default 0.5 threshold, recall for the "Died" class is only about 0.5:
the models miss about half of the patients who died. Precision is higher (about
0.7). If missing high-risk patients is more costly, the threshold can be lowered
to increase recall at the cost of precision.

## 2.4 Features selected by Lasso (combined subset)

```{r lasso-features}
lasso_comb <- glmnet_results[["Lasso Combined"]]$model
lasso_coef <- as.matrix(coef(lasso_comb$finalModel, s = lasso_comb$bestTune$lambda))
lasso_coef <- lasso_coef[rownames(lasso_coef) != "(Intercept)", 1]

selected <- lasso_coef[lasso_coef != 0]
length(selected)

# Largest coefficients (positive = higher chance of "Living", because caret
# models the second factor level, negative = higher risk of death)
top_coef <- names(sort(abs(selected), decreasing = TRUE))[1:15]
knitr::kable(tibble(feature = top_coef, coefficient = round(selected[top_coef], 3)))
```

At the selected penalty the Lasso model keeps `r length(selected)` of the
`r length(lasso_coef)` dummy-coded predictors, so it gives a much sparser model
than Ridge. The largest effects agree with clinical knowledge: more positive
lymph nodes, a higher NPI, larger tumours, older age and mastectomy (a marker of
more advanced disease) increase the risk of death, while the luminal A subtype
and a *GATA3* mutation, which is typical for luminal, hormone-sensitive tumours,
are linked to better survival. A few gene expression values (for example
*STAT5A*, *CCNB1*, *GSK3B*) are also kept.

# Part 3. Ensemble and Non-linear Models

All models in this part use the same training set, CV folds and test set as in
Part 2, so their results are directly comparable.

```{r parallel}
cl <- makeCluster(max(1, detectCores() - 1))
registerDoParallel(cl)
```

## 3.1 Hyper-parameter grids

```{r grids}
grid_knn <- expand.grid(k = seq(5, 75, by = 10))
grid_svm <- expand.grid(sigma = 2^seq(-11, -5, by = 2), C = 2^seq(-1, 5, by = 2))
tree_depths <- c(2, 3, 4, 5)
rf_depths   <- c(2, 3, 4, 5)
grid_ada <- expand.grid(n.trees = c(50, 100, 200), interaction.depth = c(1, 2, 3),
                        shrinkage = 0.1, n.minobsinnode = 10)
```

* **KNN**: number of neighbours `k`.
* **SVM (RBF kernel)**: kernel width `sigma` and cost `C`. With hundreds of
  scaled features the squared distances between patients are large, so small
  `sigma` values are needed.
* **Decision tree** and **random forest**: maximum depth 2–5.
* **AdaBoost**: implemented with `gbm` using the exponential (AdaBoost) loss;
  the grid covers the number of trees and the depth of each tree.

## 3.2 KNN and SVM

```{r knn-svm}
nonlinear_results <- list()
for (s in names(subsets)) {
  nonlinear_results[[paste("KNN", s)]] <- fit_caret_model("knn", grid_knn, subsets[[s]], "KNN", s)
  nonlinear_results[[paste("SVM", s)]] <- fit_caret_model("svmRadial", grid_svm, subsets[[s]], "SVM (RBF)", s)
}
```

## 3.3 Decision tree and random forest (depth tuned by CV)

The maximum depth is not a tuning parameter of caret's `rpart` and `ranger`
methods, so a small loop fits one model per depth and keeps the depth with the
best mean CV ROC-AUC.

```{r tree-rf}
best_by_depth <- function(depths, fit_one) {
  fits <- lapply(depths, fit_one)
  cv_auc <- sapply(fits, function(x) x$CV$ROC_AUC_mean)
  best <- fits[[which.max(cv_auc)]]
  best$bestTune <- cbind(best$bestTune, max_depth = depths[which.max(cv_auc)])
  best
}

for (s in names(subsets)) {
  nonlinear_results[[paste("Tree", s)]] <- best_by_depth(tree_depths, function(d) {
    fit_caret_model("rpart", data.frame(cp = 0.01), subsets[[s]], "Decision Tree", s,
                    control = rpart.control(maxdepth = d))
  })

  p <- length(subsets[[s]])
  nonlinear_results[[paste("RF", s)]] <- best_by_depth(rf_depths, function(d) {
    fit_caret_model("ranger",
                    expand.grid(mtry = max(1, floor(sqrt(p))), splitrule = "gini", min.node.size = 1),
                    subsets[[s]], "Random Forest", s,
                    num.trees = 500, max.depth = d, num.threads = 1)
  })
}
```

## 3.4 AdaBoost

```{r adaboost}
for (s in names(subsets)) {
  nonlinear_results[[paste("AdaBoost", s)]] <- fit_caret_model(
    "gbm", grid_ada, subsets[[s]], "AdaBoost", s,
    distribution = "adaboost", verbose = FALSE)
}

stopCluster(cl)
registerDoSEQ()
```

## 3.5 Results

```{r nonlinear-results}
nonlinear_tbl <- bind_rows(lapply(nonlinear_results, pack_row)) |> arrange(desc(CV_ROC_AUC_mean))
show_table(nonlinear_tbl)
```

```{r best-tunes}
for (nm in names(nonlinear_results)) {
  bt <- nonlinear_results[[nm]]$bestTune
  cat(sprintf("%-26s %s\n", nm, paste(names(bt), sapply(bt, as.character), sep = " = ", collapse = ", ")))
}
```

AdaBoost with the combined features has the best mean CV ROC-AUC in this part
(`r round(nonlinear_tbl$CV_ROC_AUC_mean[1], 3)`), followed by SVM, KNN and
random forest on the non-genetic features, all around 0.75. A single decision
tree is clearly the weakest model (about 0.70), which is expected because one
shallow tree has high variance and captures only a few splits.

For KNN, SVM, random forest and the decision tree, the non-genetic subset is as
good as or better than the combined one. These methods do not select features:
KNN and SVM use distances over all features, so hundreds of noisy gene
expression values dilute the useful clinical signal. Boosting, like Lasso, can
focus on a few informative features, which is why it is the only non-linear
method that gains from the combined subset.

For KNN (k = 75 on the non-genetic subset) and SVM (C = 32) the chosen values
are at the edge of the grid. A wider grid might improve these two models a
little, but their CV ROC-AUC is already on a flat plateau, so large gains are
unlikely.

# Part 4. Model Comparison and Interpretation with SHAP

## 4.1 Overall comparison

```{r comparison}
all_results <- c(glmnet_results, nonlinear_results)
all_tbl <- bind_rows(lapply(all_results, pack_row))

# Best subset for each algorithm (by mean CV ROC-AUC)
best_per_algorithm <- all_tbl |>
  group_by(Algorithm) |>
  slice_max(CV_ROC_AUC_mean, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(desc(CV_ROC_AUC_mean))

show_table(best_per_algorithm)

dir.create(file.path("..", "results"), showWarnings = FALSE)
write_csv(all_tbl |> arrange(desc(CV_ROC_AUC_mean)),
          file.path("..", "results", "model_comparison.csv"))
```

```{r cv-auc-plot, fig.height = 4}
ggplot(best_per_algorithm,
       aes(x = reorder(Algorithm, CV_ROC_AUC_mean), y = CV_ROC_AUC_mean)) +
  geom_pointrange(aes(ymin = CV_ROC_AUC_mean - CV_ROC_AUC_sd,
                      ymax = CV_ROC_AUC_mean + CV_ROC_AUC_sd)) +
  geom_point(aes(y = Test_ROC_AUC), shape = 4, size = 3, colour = "firebrick") +
  coord_flip() +
  theme_minimal() +
  labs(x = NULL, y = "ROC-AUC",
       title = "Best subset per algorithm",
       subtitle = "dot and bar: CV mean +/- SD, cross: test set")
```

```{r roc-curves, fig.height = 5}
roc_curves <- lapply(seq_len(nrow(best_per_algorithm)), function(i) {
  key <- paste(best_per_algorithm$Algorithm[i], best_per_algorithm$Subset[i])
  res <- all_results[[names(all_results)[sapply(all_results, function(r)
    paste(r$Algorithm, r$Subset) == key)]]]
  r <- roc(test_data$death_from_cancer, res$test_prob,
           levels = c("Living", "Died"), direction = "<", quiet = TRUE)
  tibble(fpr = 1 - r$specificities, tpr = r$sensitivities,
         model = sprintf("%s (%s), AUC = %.3f", res$Algorithm, res$Subset, auc(r)))
})

ggplot(bind_rows(roc_curves), aes(fpr, tpr, colour = model)) +
  geom_path() +
  geom_abline(linetype = "dashed", colour = "grey60") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.direction = "vertical") +
  labs(x = "False positive rate", y = "True positive rate", colour = NULL,
       title = "ROC curves on the test set")
```

```{r best-model}
overall_best <- best_per_algorithm |> slice(1)
overall_best

best_key <- names(all_results)[sapply(all_results, function(r)
  r$Algorithm == overall_best$Algorithm && r$Subset == overall_best$Subset)]
best_result <- all_results[[best_key]]
best_features <- subsets[[overall_best$Subset]]
```

The best model overall is **`r overall_best$Algorithm`** on the
**`r overall_best$Subset`** features, with a mean CV ROC-AUC of
`r round(overall_best$CV_ROC_AUC_mean, 3)` (SD `r round(overall_best$CV_ROC_AUC_sd, 3)`)
and a test ROC-AUC of `r round(overall_best$Test_ROC_AUC, 3)`. For most models the test
ROC-AUC is within a few hundredths of the CV estimate, so there is no sign of
overfitting to the CV folds.

The differences between the top six algorithms (CV ROC-AUC 0.749–0.765) are
smaller than one standard deviation across folds, so they perform practically
the same. In this situation the simpler and more interpretable model is
preferred, which is another reason to choose the Lasso logistic regression: it
uses only `r length(selected)` predictors and its coefficients can be read
directly. The ROC curves also show that no model dominates the others over the
whole range of thresholds.

Note: an earlier version of this analysis reached a ROC-AUC of about 0.88,
but it included `overall_survival_months` as a predictor. That variable is known
only at the end of follow-up and is strongly tied to the outcome, so the
earlier result was over-optimistic. The values reported here (about 0.77) are
a fair estimate of what the models can do with information available at
diagnosis.

## 4.2 SHAP values for the best model

SHAP values split one prediction into additive contributions of the features:
for each patient, the SHAP values of all features add up to the difference
between the predicted probability of death and the average prediction.
They are estimated here with the permutation sampling method (Štrumbelj &
Kononenko, 2014), which works for any model: features are switched one at a
time, in random order, from a background patient to the patient being explained,
and the change in the predicted probability is recorded. Averaging over many
random orders gives the SHAP value. The explanation is computed for the test
patients, with the training set as background.

```{r shap-function}
permutation_shap <- function(model, X_explain, X_background, nsim = 20, positive = "Died") {
  n <- nrow(X_explain)
  p <- ncol(X_explain)
  phi <- matrix(0, n, p, dimnames = list(NULL, names(X_explain)))
  pred_fun <- function(X) predict(model, newdata = X, type = "prob")[, positive]

  for (s in seq_len(nsim)) {
    order_s <- sample(p)                                      # random feature order
    current <- X_background[sample(nrow(X_background), n, replace = TRUE), ]
    prev <- pred_fun(current)
    for (j in order_s) {
      current[[j]] <- X_explain[[j]]                          # switch feature j to the real value
      new <- pred_fun(current)
      phi[, j] <- phi[, j] + (new - prev)                     # marginal contribution of feature j
      prev <- new
    }
  }
  phi / nsim
}
```

```{r shap-compute}
X_train_shap <- as.data.frame(train_data[, best_features])
X_test_shap  <- as.data.frame(test_data[,  best_features])

set.seed(123)
shap_time <- system.time(
  shap_values <- permutation_shap(best_result$model, X_test_shap, X_train_shap, nsim = 20)
)
shap_time[["elapsed"]]

# Check: SHAP values add up to prediction minus average prediction
pred_test <- predict(best_result$model, X_test_shap, type = "prob")[, "Died"]
base_value <- mean(predict(best_result$model, X_train_shap, type = "prob")[, "Died"])
summary(rowSums(shap_values) - (pred_test - base_value))
```

The last output checks the additivity property: for each patient the sum of
the SHAP values should equal the predicted probability minus the average
prediction (`r round(base_value, 3)`). The differences are centred at zero
(mean `r round(mean(rowSums(shap_values) - (pred_test - base_value)), 3)`); they
are not exactly zero because each patient is compared with only 20 randomly
drawn background patients. More simulations reduce this error but increase
the run time (about `r round(shap_time[["elapsed"]] / 60)` minutes here).

## 4.3 Global feature importance

```{r shap-importance, fig.height = 5}
shap_importance <- sort(colMeans(abs(shap_values)), decreasing = TRUE)
top_features <- names(shap_importance)[1:15]

tibble(feature = factor(top_features, levels = rev(top_features)),
       mean_abs_shap = shap_importance[top_features]) |>
  ggplot(aes(feature, mean_abs_shap)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(x = NULL, y = "mean |SHAP| (change in P(Died))",
       title = paste("SHAP feature importance:", overall_best$Algorithm, "-", overall_best$Subset))
```

```{r shap-summary, fig.height = 5.5}
# Colour = feature value scaled to 0-1 within each feature (factors use level order)
feature_value_01 <- function(x) {
  x <- as.numeric(x)
  if (max(x) == min(x)) return(rep(0.5, length(x)))
  (x - min(x)) / (max(x) - min(x))
}

shap_long <- bind_rows(lapply(top_features, function(f) {
  tibble(feature = f, shap = shap_values[, f], value = feature_value_01(X_test_shap[[f]]))
})) |>
  mutate(feature = factor(feature, levels = rev(top_features)))

ggplot(shap_long, aes(shap, feature, colour = value)) +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_jitter(height = 0.2, width = 0, size = 1.2, alpha = 0.8) +
  scale_colour_gradient(low = "#2c7bb6", high = "#d7191c", breaks = c(0, 1),
                        labels = c("low", "high")) +
  theme_minimal() +
  labs(x = "SHAP value (effect on P(Died))", y = NULL, colour = "Feature value",
       title = "SHAP summary plot (test set)")
```

## 4.4 Dependence plot for the most important feature

```{r shap-dependence}
top_numeric <- top_features[sapply(X_test_shap[top_features], is.numeric)][1]

tibble(value = X_test_shap[[top_numeric]], shap = shap_values[, top_numeric]) |>
  ggplot(aes(value, shap)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, colour = "firebrick") +
  theme_minimal() +
  labs(x = paste(top_numeric, "(standardised)"), y = "SHAP value",
       title = paste("SHAP dependence plot:", top_numeric))
```

## 4.5 Top 10 contributing features

```{r top10}
tibble(rank = 1:10,
       feature = names(shap_importance)[1:10],
       mean_abs_shap = round(shap_importance[1:10], 4)) |>
  knitr::kable()
```

The SHAP analysis agrees with the Lasso coefficients and with the exploratory
plots:

* **Number of positive lymph nodes** is the most important feature. Most patients
  have few positive nodes and small negative SHAP values, but patients with many
  positive nodes get a large increase in the predicted probability of death (up
  to about +0.4).
* **Age at diagnosis**, **Nottingham Prognostic Index** and **tumour size**:
  higher values increase the predicted risk.
* **Type of surgery**: mastectomy increases the predicted risk compared with
  breast-conserving surgery. This is not a causal effect of the surgery; it
  reflects that mastectomy is chosen for larger or more advanced tumours.
* **Gene expression**: high *STAT5A* expression lowers the predicted risk, while
  high *CCNB1* (a cell-cycle / proliferation gene) and *GSK3B* expression raise it.
* **GATA3 mutation** and the **PAM50 subtype** (luminal A) are linked to a lower
  risk.

The clinical features dominate the ranking, and the genetic features act as
smaller corrections. This matches the results of Parts 2 and 3, where the
genetic-only models were the weakest and adding genetic features to the
clinical ones gave only a small improvement.

# Conclusion

* After removing leakage variables (`overall_survival`, `overall_survival_months`)
  and the patient identifier, the best models reach a ROC-AUC of about 0.76–0.77
  for predicting breast-cancer–specific death.
* Clinical variables (lymph nodes, NPI, tumour size, age, receptor status and
  subtype) carry most of the predictive information. Gene expression and
  mutation data alone give ROC-AUC of about 0.70 and add only a small gain when
  combined with the clinical data.
* The penalised Lasso logistic regression on the combined features performs as
  well as the more complex non-linear models (AdaBoost, SVM, random forest) while
  being sparse and easy to interpret, so it is chosen as the final model.
* With the default threshold, recall for the "Died" class is about 0.5. In a
  clinical setting the threshold should be chosen based on the relative cost of
  missed high-risk patients and false alarms.
* Possible extensions: time-to-event models (Cox regression, random survival
  forests) that use the survival time as the outcome instead of discarding it,
  repeated CV for more stable comparisons, and threshold tuning.

```{r session-info}
sessionInfo()
```
